submission

### 2024-03-08 submission.Rmd
Max added q2

>>>>

```{r}
load("liver_data.rda")
```

### 2.1  We first want to study the gene which is the most correlated with the level of cholesterol. Identify this gene and plot `cholesterol` as a function of this gene expression level. Fit a linear model. Test if there is a significant relation between this gene expression and `cholesterol`?

```{r}
labels <- colnames(liver)

maxSlope=0
minP=1

for (i in 2:3117){
  lmgene <- lm(cholesterol~liver[,i], data=liver)
  slope <- abs(coef(lmgene, complete = TRUE)[2])
  summary_stats <- summary(lmgene)
  p_value <- summary_stats$coefficients[2, 4]

  if (slope > maxSlope){
    maxSlope=slope
    max_slope_model=lmgene
  }

  if (minP > p_value){
    minP=p_value
    min_p_model=lmgene
    genename=labels[i]
  }
}

Fitting a linear model for every gene, we get 3116 different models in total. The null-hypothesis is that a gene does no have a correlation with the cholesterol-level, in other words, the slope is zero in this case. Consequently, the gene corresponding to the model with the lowest p-value is also the model with highest slope. In our case, that turned out to be gene $A\_42P631473$ with a p-value of less than $10^{-13}$.

```{r}

ggplotRegression <- function (fit) {

require(ggplot2)

ggplot(fit$model, aes_string(x = names(fit$model)[2], y = names(fit$model)[1])) + 
  geom_point() +
  stat_smooth(method = "lm", col = "red") +
  labs(title = paste("Adj R2 = ",signif(summary(fit)$adj.r.squared, 5),
                     "Intercept =",signif(fit$coef[[1]],5 ),
                     " Slope =",signif(fit$coef[[2]], 5),
                     " P =",signif(summary(fit)$coef[2,4], 5)), x = paste(genename, "-expression"))
}

ggplotRegression(min_p_model)
#ggplotRegression(max_slope_model)

## 2.2  We now wish to perform the previous test for all of the 3116 genes. For each of the genes, fit a linear model that explains `cholesterol` as a function of the gene expression and compute the p-value of the test. AND

## 2.3  Order the p-values and plot the ordered p-values as a function of their rank. On the same plot, display the line $y = x/3116$. Discuss the result.

```{r}
labels <- colnames(liver)

p_list <- c()
lineList = c()

for (i in 2:3117){
  lmgene <- lm(cholesterol~liver[,i], data=liver)
  slope <- abs(coef(lmgene, complete = TRUE)[2])
  summary_stats <- summary(lmgene)
  p_value <- summary_stats$coefficients[2, 4]
  p_list=c(p_list, p_value)
  lineList = c(lineList, (i-1)/3116)
}

p_list=sort(p_list)
plot(p_list, col = "blue", lwd = 2, xlab = "Rank", ylab = "p-value")
lines(lineList, col = "red", lwd = 2)

legend("topleft", legend = c("p-values", "Uniform line"), col = c("blue", "red"), lwd = 2)

```
Plotting all values, we can conclude that there are amount of models that have a p-value below 0.05. It is tempting to make the conclusion that all these models are statistically significant since they all have a small p-value. However, in the event that the we only have random data, we still expect 5\% of the models to appear significantly significant. In fact, if we were to only have random data, we expect a uniform distribution between 0 and 1 for the p-values. Visually, looking at the Figure above, we can the p-values perform better that the uniform distribution and therefore appear to contain significant findings.

```{r}
labels <- colnames(liver)

p_list_shuffeled <- c()
lineList_shuffeled = c()

for (i in 2:3117){
  cholesterol=liver$cholesterol
  randomLabel <- sample(cholesterol)
  lmgene <- lm(randomLabel~liver[,i], data=liver)
  slope <- abs(coef(lmgene, complete = TRUE)[2])
  summary_stats <- summary(lmgene)
  p_value <- summary_stats$coefficients[2, 4]
  p_list_shuffeled=c(p_list_shuffeled, p_value)
  lineList_shuffeled = c(lineList_shuffeled, (i-1)/3116)
}

p_list_shuffeled=sort(p_list_shuffeled)
plot(p_list_shuffeled, col = "blue", lwd = 2, xlab = "Rank", ylab = "p-value", main = "Shuffeled labels")
lines(lineList_shuffeled, col = "red", lwd = 2)

legend("topleft", legend = c("p-values when shuffeled labels", "Uniform line"), col = c("blue", "red"), lwd = 2)

```
 In the nest question, we are tasked with finding the maximum number of genes such that the expected number of false discoveries is at most 5\%. For this method, we use the Benjamini-Hochberg procedure \cite{FDR}. In short, it states that if we have a set of $\{ H_1, H_2, ... \: , H_m \}$ with associated and ordered p-values $\{ p_1, p_2, ... \: , p_m \}$, then the threshold for the index can be written as:

##2.4

$$ P_{BH}= max_i \left\{ P_{(i)} : P_{(i)} \leq \alpha \frac{i}{m}      \right\} $$

```{r}

soft_guarantee=c()
alpha=0.05
index=0
n=3116

for (i in 1:3116){
  val=i*alpha/n
  soft_guarantee=c(soft_guarantee, val)
  if (p_list[i]<val){
    index=i
  }

print(index)

plot(p_list, col = "blue", lwd = 2, xlab = "Index", ylab = "p-value")
lines(soft_guarantee, col = "red", lwd = 2)
abline(v=index, col="black")
legend("topleft", legend = c("Benjamini-Hochberg boundary", "P-value", paste("Intersect=", as.character(index))), col = c("red", "blue", "black"), lwd = 2)

plot(p_list[1:1200], col = "blue", lwd = 2, xlab = "Index", ylab = "p-value")
lines(soft_guarantee[1:1200], col = "red", lwd = 2)
abline(v=index, col="black")
legend("topleft", legend = c("Benjamini-Hochberg boundary", "P-value", paste("Intersect=", as.character(index))), col = c("red", "blue", "black"), lwd = 2)

```
We can see that $\alpha \frac{i}{m}$ plots out a line, and we choose i such that $P_{(i)}$ always is smaller or equal to that value. Doing so, we get 941 discoveries such that the expected number of false discoveries is no more than 5 \%

##2.5 A stronger guarantee would be that on impose that there shouldn't be more than a 5 \% risk that any discovery is false. For this method, we use the Holm–Bonferroni method. The method is as follows:\\

I have my ordered p-values, $\{ P_1, P_2, ... \: , P_m \}$, and want the FWER to be no higher than a level $\alpha$.
Is $P_1 < \alpha/m$? so, reject $H_1$ and continue to the next step, otherwise EXIT.
Is $P_2 < \alpha/(m-1)$? so, reject $H_2$ and continue to the next step, otherwise EXIT.

Continue this iterative process as long as $P_k < \frac{\alpha}{m+1-k}$. Using this process, we get the following result:

```{r}

strong_guarantee=c()
alpha=0.05
index=0

for (i in 1:n){
  val=alpha/(n+1-i)
  strong_guarantee=c(strong_guarantee, val)
  if (p_list[i]<val){
    index=i
  }

plot(p_list, col = "blue", lwd = 2, xlab = "Index", ylab = "p-value")
lines(strong_guarantee, col = "red", lwd = 2)
abline(v=index, col="black", )
legend("topleft", legend = c("Holm–Bonferroni", "P-value", paste("Intersect=", as.character(index))), col = c("red", "blue", "black"), lwd = 2)

plot(p_list[1:225], col = "blue", lwd = 2, xlab = "Index", ylab = "p-value")
lines(strong_guarantee, col = "red", lwd = 2)
abline(v=index, col="black", )
legend("bottomright", legend = c("Holm–Bonferroni", "P-value", paste("Intersect=", as.character(index))), col = c("red", "blue", "black"), lwd = 2)

Using this process, we see that we get 223 discoveries. We also not that smallest p-value in that set is approximately equal to $1.7 \cdot 10^{-5}$.
